\documentclass[12pt]{article}
\usepackage{fullpage}
\setlength{\parskip}{.8ex}
\setlength{\parindent}{0ex}
\usepackage{url}
\usepackage{xspace} %
\usepackage{alltt} %
\usepackage[T1]{fontenc} %
\usepackage{graphicx}

\usepackage{tikz}                %
\usetikzlibrary{arrows,positioning} 
\tikzset{
    %
    >=stealth',
    %
    absbox/.style={
           rectangle, rounded corners, draw=black, very thick,
           text width=14em, minimum height=2em,
           text centered},
    %
    absarrow/.style={
           <-, thick,
           shorten <=2pt,
           shorten >=2pt},
    field/.style={ %
           rectangle, draw=black, thick,
           text width=4.2em, minimum height=1.6em,
           text centered, inner sep=0pt},
    cell/.style={ %
           rectangle, draw=black, thick,
           text width=1.5em, minimum height=1.5em,
           text centered, inner sep=0pt},
    pointer/.style={ %
           <-, thick,
           shorten <=2pt,
           shorten >=2pt},
    vertex/.style={ %
           circle, draw=black, thick,
           text width=1.5em, minimum height=1.5em,
           text centered, inner sep=0pt},
}

\newcommand{\notes}[1]{} %
\newcommand{\todo}[1]{}
 %
\newcommand{\body}[1]{} %

\newcommand{\mysubsec}[1]{\subsection*{#1}}
\newcommand{\mypar}[1]{\vspace{2ex}\noindent\b{#1.}~}

 %

 %
 %





\hyphenation{in-cre-men-tal-i-za-tion di-men-sion-al}\todo{}
\def\hyph{-\penalty0\hskip0pt\relax}

\newcommand{\defn}[1]{{\it #1}\index{#1|bold}}
 %
 %
 %


 %
 %

\newcommand{\lang}[1]{#1\index{#1}}

\def\mathify#1{\ifmmode{#1}\else\mbox{$#1$}\fi}
\newcommand{\m}[1]{$#1$} %

\newenvironment{code}{\begin{center}\small\begin{alltt}\vspace{-1.5ex}}{\vspace{-1.5ex}\end{alltt}\end{center}}
\renewcommand{\c}[1]{{\tt\small #1}} %
 %
\renewcommand{\b}[1]{{\bf #1}} %





 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
 %
\newcommand{\p}[1]{{\small\mathify{\it #1}}}
 %
 %
\renewcommand{\to}{\c{->}\xspace} %
\renewcommand{\|}{\mathify{\,\mbox{\c{|}}\,}} %

 %
\renewcommand{\=}{\mathify{\,=\,}\linebreak[0]}




\newcommand{\s}[1]{\mathify{\#}\c{#1}}





 %
 %






 %

 %
 %




 %






 %
 %
 %
 %
 %


















\begin{document}

\body{} %

\newpage
\appendix

\renewcommand{\thesection}{\arabic{section}}

\setcounter{page}{1}
\setcounter{section}{0}


\notes{}

\section*{{\Large DistAlgo Language Description}}

{\large Yanhong A.\ Liu, Bo Lin, and Scott Stoller}

\c{liu@cs.stonybrook.edu, bolin@cs.stonybrook.edu, stoller@cs.stonybrook.edu}

{\small Revised March 24, 2017}
\vspace{4ex}

\def\langmap{\vspace{.5ex}\m{\longrightarrow}~\b{in Python syntax:~}
  \vspace{.0ex}}
 %
\renewcommand{\mysubsec}[1]{\section{#1}}
\renewcommand{\mypar}[1]{\vspace{-1ex}\subsection{#1}\vspace{-1ex}}
\newcommand{\mysubpar}[1]{\vspace{-2ex}\paragraph{#1.}}


DistAlgo is a language for distributed algorithms.  We describe DistAlgo
language constructs as extensions to conventional object-oriented
programming languages, including a syntax for extensions to Python.

There are four components conceptually: (1) distributed processes and
sending messages, (2) control flows and receiving messages, (3) high-level
queries of message histories, and (4) configurations.

High-level queries are not specific to distributed algorithms, but using
them over message histories is particularly helpful for expressing and
understanding distributed algorithms at a high level.  Some conventional
programming languages, such as Python, support high-level queries to some
extent, but DistAlgo query constructs are more declarative, especially with
the support of tuple patterns for messages.


\mysubsec{Distributed processes and sending messages}

\mypar{Process definition}

A process definition is of the following form.  It defines a type of
processes named \p{p}, by defining a class \p{p} that extends class
\c{process}.  The \c{\p{process\_body}} is a set of method definitions and
handler definitions, to be described.
\begin{code}
    class \p{p} extends process:
      \p{process\_body}
\end{code}
The syntax of process definition could be made simpler and clearer:
\begin{code}
    process \p{p}:
      \p{process_body}
\end{code}
but it would make \c{process} a keyword, which is usually a reserved word,
whereas \c{process} as a class name is not reserved %
and can be defined or redefined to be anything else.

\langmap
\begin{code}
    class \p{p} (process):
      \p{process_body}
\end{code}

\todo{}


A special method \c{setup} may be defined in \c{\p{process\_body}} for
initially setting up data in the process before the process's execution
starts.
For each parameter \p{v} of \c{setup}, a process field named \p{v} is
defined automatically and assigned the value of parameter \p{v}; additional
fields
can be defined explicitly in the method body of \c{setup}.

A special method \c{run()}
may be defined in \c{\p{process\_body}} for carrying out the main flow of
execution.


A special variable \c{self} refers to the process itself.  
Fields of the process may be defined by including the field name
as a parameter of method \c{setup}, or by explicitly prefixing the field name 
with \c{self} in an assignment to the field.
References to fields of the process 
do not need to be prefixed with \c{self}.  References to methods of the
process do not need to be prefixed with \c{self} either.  Also, method
definitions implicitly include parameter \c{self}.




\mypar{Process creation}

Process creation consists of statements for creating, setting up, and
starting processes. %

A process creation statement is of the following form.  It creates \p{n}
new processes of type \p{p} at each node in the value of \c{\p{node\_exp}},
and returns\notes{} the resulting process
or set of processes\notes{}. %
\begin{code}
    \p{n} new \p{p} at \p{node_exp}
\end{code}
Expression \c{\p{node\_exp}} evaluates to a node
or a set of nodes, specifying where the new processes will be created.
A node is a running DistAlgo program on a machine. %
A node is identified by a string of the form \c{name@host}, where \c{name}
can be specified on the command line when starting the node, and \c{host}
is the host name of the machine running the node; \c{@host} can be omitted
if the node is running on the local machine.
All nodes communicating with each other must have the same cookie, which
can be specified on the command line when starting the node.
The number \p{n} and the \c{at} clause are optional; the defaults are 1 and
local node, respectively.  When both are omitted, a single process is
created and returned\notes{}; otherwise, a set of
processes is created and returned\notes{}.

\langmap
\begin{code}
    new(\p{p}, num = \p{n}, at = \p{node_exp})
\end{code}



A process setup statement is of the following form.  It sets up the process
or set of processes that is the value of expression \p{pexp}, using method
\c{setup} of the process or processes with the values of argument
expressions \p{args}.  If the values of \p{args} are available when the
process or processes are created at a call to \c{new}, the call to
\c{setup} can be omitted by inserting tuple \c{(\p{args})} after \p{p} in
the call to \c{new}.
\begin{code}
    \p{pexp}.setup(\p{args})
\end{code}

\langmap%
\begin{code}
    setup(\p{pexp}, (\p{args}))        \normalfont{Note: You must add a trailing comma if \p{args} is a single argument.}
\end{code}

A process start statement is of the following form.  It starts the
execution of the method \c{run} of the process or set of processes that is
the value of expression \p{pexp}.
\begin{code}
    \p{pexp}.start()
\end{code}

\langmap%
\begin{code}
    start(\p{pexp})
\end{code}

\notes{}

\mypar{Sending messages}

A statement for sending messages is of the following form.  It sends the
message that is value of expression \p{mexp} to the process or set of
processes that is the value of expression \p{pexp}.  A message can be any
value but is by convention a tuple whose first component is a string,
called a tag, indicating the kind of the message.
\begin{code}
    send \p{mexp} to \p{pexp}
\end{code}

\langmap
\begin{code}
    send(\p{mexp}, to = \p{pexp})
\end{code}
\notes{}


\mysubsec{Control flows and receiving messages}

Received messages can be handled both asynchronously, using yield points
and \c{receive} definitions, and synchronously, using \c{await} statements.


\mypar{Yield points}

A yield point preceding a statement is of the following form, where
identifier \p{l} is a label and is optional.  It specifies that point in
the program as a place where control yields to handling of un-handled
messages, if any, and resumes afterwards.
\begin{code}
    -- \p{l}:
\end{code}

\langmap
\begin{code}
    -- \p{l}
\end{code}
\quad\quad which is a statement in Python, where \p{l} is any valid Python
identifier.
\notes{}

\notes{}


\mypar{Handling received messages}


A handler definition, also called a \c{receive} definition, is of the
following form.
It handles, at yield points labeled \p{l\sb{1}}, ..., \p{l\sb{j}},
un-handled messages that match \p{mexp} %
from \p{pexp}, where \p{mexp} and \p{pexp} are parts of a pattern,
to be described.
The \c{from} and \c{at} clauses are optional; the defaults are any process
and all yield points.  The \c{\p{handler\_body}} is a sequence of
statements to be executed for the matched messages.
\begin{code}
    receive \p{mexp} from \p{pexp} at \p{l\sb{1}}, ..., \p{l\sb{j}}:
      \p{handler_body}
\end{code}

\langmap
\begin{code}
    def receive(msg = \p{mexp}, from_ = \p{pexp}, at = (\p{l\sb{1}}, ..., \p{l\sb{j}})):
      \p{handler_body}
\end{code}
\quad\quad where \c{\_} is added after \c{from} because \c{from} is a
reserved word in Python.


\mypar{Synchronization}


A simple \c{await} statement is of the following form.  It waits for the
value of Boolean-valued expression \p{bexp} to become true.
\begin{code}
    await \p{bexp}
\end{code}

\langmap
\begin{code}
    await(\p{bexp})
\end{code}

A general, nondeterministic \c{await} statement is of the following form.
It waits for any of the values of expressions \p{bexp\sb{1}}, ...,
\p{bexp\sb{k}} to become true or a timeout after \p{t} seconds, 
and then nondeterministically selects one\notes{} of statements
\p{stmt\sb{1}}, ..., \p{stmt\sb{k}}, \p{stmt} whose corresponding
conditions are satisfied to execute.  The \c{or} and \c{timeout} clauses
are optional.
\begin{code}
    await \p{bexp\sb{1}}: \p{stmt\sb{1}}
    or ... 
    or \p{bexp\sb{k}}: \p{stmt\sb{k}}
    timeout \p{t}: \p{stmt}
\end{code}
\notes{}

\langmap
\begin{code}
    if await(\p{bexp\sb{1}}): \p{stmt\sb{1}}
    elif ...
    elif \p{bexp\sb{k}}: \p{stmt\sb{k}}
    elif timeout(\p{t}): \p{stmt}
\end{code}

An \c{await} statement must be preceded by a yield point, for handling
messages while waiting; if a yield point is not specified explicitly, the
default is that all message handlers can be executed at this point.

\notes{}


\mysubsec{High-level queries of message histories}

High-level queries can be used over message histories, and patterns
can be used for matching the messages.


\mypar{Message histories}
Histories of message sent and received by a process are kept in variables
\c{sent} and \c{received}, respectively.
Sequence \c{sent} is updated at each \c{send} statement, by adding each
message sent to a process.
Sequence \c{received} is updated at the next yield point if there are
un-handled messages, by adding un-handled messages before any matching
receive handler executes.

The following two expressions are equivalent and return true iff a message
that matches \p{mexp} to \p{pexp} is in \c{sent}.  The \c{to} clause is
optional; the default is any process.
\begin{code}
    sent \p{mexp} to \p{pexp}
    \p{mexp} to \p{pexp} in sent
\end{code}

\langmap
\begin{code}
    sent(\p{mexp}, to = \p{pexp})
    (\p{mexp}, \p{pexp}) in sent
\end{code}

Similarly, the following expressions use \c{received}.
\begin{code}
    received \p{mexp} from \p{pexp}
    \p{mexp} from \p{pexp} in received
\end{code}

\langmap
\begin{code}
    received(\p{mexp}, from_ = \p{pexp})
    (\p{mexp}, \p{pexp}) in received
\end{code}
\todo{}


\mypar{Patterns}
A pattern can be used to match a message, in \c{sent} and \c{received}, and
by a \c{receive} definition.  A constant value, such as \c{'ack'}, or a
previously bound variable, indicated with prefix \c{=}, in the pattern must
match the corresponding components of the message.  An underscore (\c{\_})
matches anything.
Previously unbound variables in a pattern are bound to the corresponding
components in the matched message. %
In general, any data construction expression %
can be used as a pattern; we use only tuple patterns because messages are
by convention tuples.

For example, \c{received ('ack',=n,\_) from a} matches every triple in
\c{received} whose first two components are \c{'ack'} and the value of
\c{n}, and binds \c{a} to the sender.

Notation \c{=x} means a value that is equal to the value of variable \c{x};
it is equivalent to using a fresh variable \c{y} in its place and adding a
test \c{y=x} before \c{x} is used.
This notation can generalize: one can add as prefix any binary operator
that is a symbol not allowed in identifiers, uses the parameter value as
the right operand, and returns a Boolean value.  For example, \c{>x} means
a value that is greater than the value of parameter \c{x}.

\langmap
\c{\_} is used in place of \c{=} to indicate previously bound variables.
This forbids the use of variable names that start with \c{\_} in the query.
Other binary operators cannot be used as prefix.


\mypar{Comprehension, aggregation, and quantification}
A query is over sets or sequences.  It can be a comprehension, aggregation,
or quantification, plus a set of \defn{parameters}---variables whose values
are bound before the query.
For a query to be well-formed, every variable in it must be
\defn{reachable} from a parameter---be a parameter or be the left-side
variable of a membership clause whose right-side variables are reachable.

To indicate that a variable \c{x} on the left side of a membership
clause is a parameter, add prefix \c{=} to \c{x}; this is only needed
for the first occurrence of such a variable.
Use of \c{=} to indicate parameters is consistent with use of \c{=} in
patterns.

\pagebreak
\mysubpar{Comprehension}
A comprehension is a query of the following form.  Given values of
parameters, the comprehension returns the set of values of \p{exp} for all
combinations of values of variables that satisfy all membership clauses
\c{\p{v_i} in \p{sexp\sb{i}}} and condition \p{bexp}.
\begin{code}
    \{\p{exp}: \p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}}, \p{bexp}\}
\end{code}

\langmap
\begin{code}
    setof(\p{exp}, \p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}}, \p{bexp})
\end{code}





\mysubpar{Aggregation}
An aggregation is a query of one of the following two forms, where
\c{\p{agg}} is an aggregation operator, including \c{count}, \c{sum},
\c{min}, and \c{max}.
Given values of parameters, the aggregation returns the value of applying
\c{\p{agg}} to the set value of \c{\p{sexp}}, for the first form, or to the
multiset of values of \p{exp} for all combinations of values of variables
that satisfy all membership clauses \c{\p{v\sb{i}} in \p{sexp\sb{i}}} and
condition \p{bexp}, for the second form.
\begin{code}
    \p{agg} \p{sexp}
    \p{agg} (\p{exp}: \p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}}, \p{bexp})
\end{code}

\langmap
\begin{code}
    \p{agg}(\p{sexp})
    \p{agg}of(\p{exp}, \p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}}, \p{bexp})
\end{code}
\quad\quad where \c{len} is used in place of \c{count}.


\mysubpar{Quantification}
A quantification is a query of one of the following two forms, called
existential and universal quantifications, respectively.
Given values of parameters, the quantification returns \c{true} iff for
some or all, respectively, combinations of values of variables that satisfy
all membership clauses \c{\p{v\sb{i}} in \p{sexp\sb{i}}},
condition %
\p{bexp} evaluates to \c{true}.
When an existential quantification returns \c{true}, all variables in
the query are also bound to a combination of values, called a witness,
that satisfy all the membership clauses and condition \p{bexp}.
\begin{code}
    some \p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}} has \p{bexp}
    each \p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}} has \p{bexp}
\end{code}

\langmap\notes{}
\begin{code}
    some(\p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}}, has = \p{bexp})
    each(\p{v\sb{1}} in \p{sexp\sb{1}}, ..., \p{v\sb{k}} in \p{sexp\sb{k}}, has = \p{bexp})
\end{code}
\notes{}

\mysubpar{Other forms for membership and condition}
In all of comprehension, aggregation, and quantification, 
in membership clause \c{\p{v\sb{i}} in \p{sexp\sb{i}}},
if \p{sexp_i} is a variable \p{s_i}, then clause \c{\p{v_i} in \p{s_i}} can
also be written as \p{s_i(v_i)}.
Also, 
a tuple pattern may be used in place of variable \p{v_i}.
When condition \p{bexp} is \c{true}, the clause with \p{bexp} can be
omitted.

\langmap
Only for \p{sexp_i} being variable \c{sent} or \c{received} can clause
\c{\p{v_i} in received} or \c{\p{v_i} in sent} be written as
\c{received(\p{v_i})} or \c{sent(\p{v_i})}, respectively.



\mysubsec{Configurations}



\mypar{Channel types}

The following statement configures all channels to be
first-in-first-out (FIFO).
Other options for \c{channel} include \c{reliable} and \c{\{reliable,
  fifo\}}.  When these options are specified, TCP is used for process
communication; otherwise, UDP is used.
\begin{code}
    configure channel = fifo
\end{code}

\langmap
\begin{code}
    config(channel = 'fifo')
\end{code}
Channels can also be configured separately for messages from certain types
of processes to certain types of processes, by adding clauses \c{from
  \p{ps}} and \c{to \p{qs}}, or arguments \c{from\_ = \p{ps}} and \c{to =
  \p{qs}} in Python syntax, where \p{ps} and \p{qs} can be a type of
processes or a set of types of processes.  Each of these clauses is
optional; the default is all types of processes.


\mypar{Message handling}

The following statement configures the system to handle all un-handled
messages at each yield point; this is the default.
Other options for \c{handling} include \c{one}.
\begin{code}
    configure handling = all
\end{code}

\langmap
\begin{code}
    config(handling = 'all')
\end{code}


\mypar{Logical clocks}

The following statement configures the system to use Lamport clock.  Other
options for \c{clock} include \c{vector}; it is currently not supported.
\begin{code}
    configure clock = Lamport
\end{code}

\langmap
\begin{code}
    config(clock = 'Lamport')  
\end{code}

A call \c{logical\_time()} returns the current value of the logical
clock.


\mypar{Overall}

A DistAlgo program is written in files named with extension \c{.da}.  It
consists of a set of process definitions, a method \c{main}, and possibly
other, conventional program parts.  Method \c{main} specifies the
configurations and creates, sets up, and starts a set of processes.

DistAlgo language constructs can be used in process definitions and method
\c{main} and are implemented according to the semantics described; 
other, conventional program parts are implemented according to their
conventional semantics.
\notes{}



\mysubsec{Other useful functions in Python}

\mypar{Logging output}

The following method prints the values of expressions \c{\p{exp_1}} through
\c{\p{exp_k}} in their \c{str()} representation, separated by the value of
\c{\p{str\_exp}} and prefixed with system timestamp, process id, and the
specified integer level \p{l}, to the log of %
the node that runs the current DistAlgo process; the printing is done only
if level \p{l} is greater or equal to the default logging level or the
level specified on the command line when starting the node. %
The log defaults to console\notes{}\notes{}, but can be a file specified on the command
line when starting the node. %
\begin{code}
    output(\p{exp\sb{1}}, ..., \p{exp\sb{k}}, sep = \p{str\_exp}, level = \p{l})
\end{code}
Argument \c{sep} is optional and defaults to the empty space.  Argument
\c{level} is optional and defaults to \c{logging.INFO}, corresponding to
value 20, in the Python logging module; see
\url{https://docs.python.org/3/library/logging.html#levels} for a list of
predefined level names.


\mypar{Importing modules}

A DistAlgo module \p{module} is a DistAlgo file 
named \c{\p{module}.da}.  DistAlgo files can be imported just as Python
files.

For example, the following statement takes DistAlgo module \p{module} in
file \c{\p{module}.da}, compiles it if an up-to-date compiled file does not
already exist, and assigns to \p{m} the resulting module object if
successful or raises \c{ImportError} otherwise.
\begin{code}
  import \p{module} as \p{m}
\end{code}


\end{document}
